---
title: "Response to Reviewer comments"
author: ""
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    code_folding: hide
---
# Load Libraries

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE, warning = FALSE, message = FALSE,
    fig.width = 10, fig.height = 7, dpi = 300
)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

library(tidyverse)
library(readxl)
library(lme4)
library(lmerTest)
library(ggpubr)
library(patchwork)
library(car)
library(emmeans)
library(ggplot2)
library(vegan)
```

# Load Metadata

```{r metadata}
TSF <- readRDS("Processed_data/metadata/fire/Fire_regime_MER.Rdata") %>%
    distinct(Site, severity, TSF_years)

MER_Site_description <- read_excel("Processed_data/metadata/veg/MER_Site_description.xlsx",
    sheet = "SP suggestions (highlighted)"
)

MER_aridity <- readRDS("Processed_data/metadata/Aridity_MER.Rdata")

Mer_veg <- MER_Site_description %>%
    left_join(TSF) %>%
    dplyr::select(Site, Site_nickname, Last_fire_severity, severity, TSF_years,
        Veg_type, Veg_type_alt, Veg_type_alt_2, Fire_interval,
        Min_TFI = `Min TFI`,
        Simple_fire_response_dominants, Fire_response
    ) %>%
    mutate(Min_TFI = as.numeric(Min_TFI)) %>%
    left_join(MER_aridity)

# read in host frequency d   ta
myco_freq_plot <- readRDS("Processed_data/metadata/veg/myco_freq_plot_level_df.Rdata") %>%
    select(Plot, freq_AM, freq_ECM)
```

# Load Guild Relative Abundance Data

```{r load-data}
dat_hyph <- readRDS("Processed_data/Seq_dat/Hyph/all_guild_RA_Hyph.rds")
dat_soil <- readRDS("Processed_data/Seq_dat/Soil/all_guild_RA_soil.rds")
```

```{r inspect-data}
cat("--- Hyph (Myc) columns ---\n")
colnames(dat_hyph)

cat("\n--- Soil ITS columns ---\n")
colnames(dat_soil)

cat("\n--- Hyph guilds ---\n")
unique(dat_hyph$guild2)

cat("\n--- Soil guilds ---\n")
unique(dat_soil$guild2)
```

# Join Metadata

```{r join-metadata}
# Summarise guild RA per plot for Hyph
# hyph_guild_plot <- dat_hyph %>%
#     group_by(Site, Plot, Fire_Treatment) %>%
#     summarise(
#         guild_reads = sum(count, na.rm = TRUE),
#         .groups = "drop"
#     ) %>%
#     ungroup() %>%
#     left_join(Mer_veg, by = "Site")%>%
#   mutate(
#     excluded = myco_reads == 0,
#     Fire_Treatment = as.factor(Fire_Treatment),
#     Site = as.factor(Site)
#   ) %>%
#   left_join(myco_freq_plot, by = "Plot") %>%
#   rename(
#     host_freq_AM = freq_AM,
#     host_freq_EcM = freq_ECM,
#     fire_severity = severity,
#   )

# Summarise guild RA per plot for Soil
soil_guild_plot <- dat_soil %>%
    group_by(Site, Plot, Fire_Treatment) %>%
    summarise(
        total_reads = sum(count, na.rm = TRUE),
        myco_reads = sum(count[str_detect(guild2, "(?i)mycorrhizal")], na.rm = TRUE),
        Unassigned_reads = sum(count[str_detect(guild2, "(?i)Unassigned")], na.rm = TRUE),
        .groups = "drop"
    ) %>%
    left_join(Mer_veg, by = "Site") %>%
    mutate(
        excluded = myco_reads == 0,
        Fire_Treatment = as.factor(Fire_Treatment),
        Site = as.factor(Site)
    ) %>%
    left_join(myco_freq_plot, by = "Plot") %>%
    rename(
        host_freq_AM = freq_AM,
        host_freq_EcM = freq_ECM,
        fire_severity = severity,
    )
```

```{r check-join}
# cat(
#     "Hyph plots with severity info:",
#     sum(!is.na(hyph_guild_plot$fire_severity)), "/", nrow(hyph_guild_plot), "\n"
# )
cat(
    "Soil plots with severity info:",
    sum(!is.na(soil_guild_plot$fire_severity)), "/", nrow(soil_guild_plot), "\n"
)
```

# Reviewer 2 Comments {.tabset}

## Comment 1

```{r rev2-comment1}
# ============================================================
# Test whether excluded samples (0 mycorrhizal reads) are
# biased toward high fire severity, particular sites, or
# whether exclusion is better explained by technical factors
# (low read depth) or ecological context (low host frequency)
# ============================================================
# Verify structure
str(soil_guild_plot)
table(soil_guild_plot$excluded) # how many excluded vs retained?

ex_samps <- soil_guild_plot %>%
    filter(excluded == TRUE) %>%
    count(Site, Fire_Treatment, fire_severity, total_reads, host_freq_AM, host_freq_EcM, myco_reads, Unassigned_reads)

print(ex_samps, n = Inf)

# ── 2. Logistic mixed model: is exclusion predicted by
#       fire severity or site, after accounting for
#       read depth and host frequency? ─────────────────────
m_exclusion <- glmer(
    excluded ~ fire_severity +
        scale(total_reads) +
        scale(host_freq_AM) +
        scale(host_freq_EcM) +
        (1 | Site),
    data = soil_guild_plot,
    family = binomial(link = "logit")
)

summary(m_exclusion)
Anova(m_exclusion, type = "III") # Type III LRT for each predictor

# ── 3. Test site-level bias directly ────────────────────────
# Chi-square test: are excluded samples distributed evenly
# across sites?
site_exclusion_table <- table(soil_guild_plot$Site, soil_guild_plot$excluded)
chisq.test(site_exclusion_table)

# Proportions excluded per site (for reporting / plotting)
exclusion_by_site <- as.data.frame(prop.table(site_exclusion_table, margin = 1))
colnames(exclusion_by_site) <- c("site", "excluded", "proportion")
exclusion_by_site[exclusion_by_site$excluded == TRUE, ]

# ── 4. Test fire severity bias directly ─────────────────────
# Among burnt plots only: are excluded samples concentrated
# in high-severity burns?
burnt_only <- soil_guild_plot[soil_guild_plot$Fire_Treatment == "B", ]

m_severity_only <- glmer(
    excluded ~ scale(fire_severity) +
        scale(total_reads) +
        (1 | Site),
    data = burnt_only,
    family = binomial(link = "logit")
)

summary(m_severity_only)
Anova(m_severity_only, type = "III")

# ── 5. Visualise exclusion rate by fire severity class ──────
ggplot(soil_guild_plot, aes(x = factor(fire_severity), fill = factor(excluded))) +
    geom_bar(position = "fill") +
    scale_fill_manual(
        values = c("grey70", "firebrick"),
        labels = c("Retained", "Excluded")
    ) +
    labs(
        x = "Fire severity class (Chafer et al. 2004)",
        y = "Proportion of samples",
        fill = "Sample status",
        title = "Exclusion rate by fire severity"
    ) +
    theme_classic()

# ── 6. Visualise exclusion rate by site ─────────────────────
ggplot(soil_guild_plot, aes(x = Site, fill = factor(excluded))) +
    geom_bar(position = "fill") +
    scale_fill_manual(
        values = c("grey70", "firebrick"),
        labels = c("Retained", "Excluded")
    ) +
    labs(
        x = "Site",
        y = "Proportion of samples",
        fill = "Sample status",
        title = "Exclusion rate by site"
    ) +
    theme_classic() +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))

# ── 7. Summary table for reporting ──────────────────────────
exclusion_summary <- aggregate(
    excluded ~ Site + Fire_Treatment,
    data = soil_guild_plot,
    FUN = function(x) {
        c(
            n_excluded = sum(x), n_total = length(x),
            prop = mean(x)
        )
    }
)
print(exclusion_summary)
```



## Comment 2

```{r}
# From ITS prep script
wide_myco_soil <- readRDS("Processed_data/Seq_dat/Soil/wide_myco.rds")
myco_tax_Soil <- readRDS("Processed_data/Seq_dat/Soil/myco_tax_Soil.rds")
All_Sites <- readRDS("raw_data/MER_Site_Data/All_Sites.RDS") %>%
    select(Site, Plot, Fire_Treatment)
MER_fire_regime <- readRDS("Processed_data/metadata/fire/Fire_regime_plot.rds")

# first remove samples that have no mycorrhizal reads because they cause errors below
wide_myco_soil <- All_Sites %>%
    left_join(wide_myco_soil) %>%
    filter(!if_all(starts_with("ITSall"), ~ . == 0))

# ── 1. Add fire severity and TSF to metadata for plotting ───
wide_myco_soil <- wide_myco_soil %>%
    left_join(
        All_Sites %>%
            left_join(
                MER_fire_regime %>%
                    select(
                        Site, Plot, Fire_Treatment, severity, fire_severity_num, fire_interval
                    )
            )
    )

# extract the community table, save as a new object
mat_myco <- wide_myco_soil %>% select(starts_with("ITSall"))

# ── 2. PERMANOVA: test fire severity and TSF on composition ─
# Model 1: Fire treatment + Site (your existing model)
perm_base <- adonis2(
    mat_myco ~ Fire_Treatment + Site,
    data     = wide_myco_soil,
    distance = "robust.aitchison",
    by       = "margin"
)
print(perm_base)

perm_base <- adonis2(
    mat_myco ~ fire_severity_num + Site,
    data     = wide_myco_soil,
    distance = "robust.aitchison",
    by       = "margin"
)
print(perm_base)

# Model 2: Add fire severity and TSF within burnt plots only
burnt_myco <- wide_myco_soil %>% filter(Fire_Treatment == "B")
mat_myco_burnt <- burnt_myco %>% select(starts_with("ITSall"))

perm_burnt <- adonis2(
    mat_myco_burnt ~ fire_severity + Fire_interval + TSF_years + Veg_type,
    data     = burnt_myco,
    distance = "robust.aitchison",
    by       = "margin"
)
print(perm_burnt)

cap_site <- capscale(mat_myco ~ Fire_Treatment + Site, data = wide_myco_soil, distance = "robust.aitchison", add = TRUE)
anova(cap_site)
summary(cap_site)
Cap1_aov <- as.data.frame(anova(cap_site, by = "margin")) %>%
    rownames_to_column()
cap_site
proportions <- round(cap_site$CCA$eig / cap_site$tot.chi * 100, 1) # proportion of variation associated with each axis


cap_burnt <- capscale(mat_myco_burnt ~ fire_severity + Fire_interval + TSF_years + Veg_type, data = burnt_myco, distance = "robust.aitchison", add = TRUE)
anova(cap_burnt)
summary(cap_burnt)
Cap1_aov <- as.data.frame(anova(cap_burnt, by = "margin")) %>%
    rownames_to_column()
cap_burnt
proportions_burnt <- round(cap_burnt$CCA$eig / cap_burnt$tot.chi * 100, 1) # proportion of variation associated with each axis


# produce a nice plot
# first extract scores from the resulting object and subset out different types of scores
scrs <- scores(cap_site, tidy = TRUE)
scrs_spp <- scrs %>% filter(score == "species")
scrs_site <- scrs %>% filter(score == "sites")
scrs_cent <- scrs %>% filter(score == "centroids")
scrs_biplot <- scrs %>% filter(score == "biplot")

# Extract PERMANOVA stats for annotation
perm_fire <- perm_base["Fire_Treatment", ]
perm_site <- perm_base["Site", ]
perm_df_res <- attr(perm_base, "heading") # residual df is in perm_base
perm_res_df <- perm_base["Residual", "Df"]

make_perm_label <- function(perm_tbl, res_df) {
    fire <- perm_tbl["Fire_Treatment", ]
    site <- perm_tbl["Site", ]
    paste0(
        "PERMANOVA:\n",
        "Fire  F(1,", res_df, ") = ", round(fire$F, 2),
        ", R\u00b2 = ", round(fire$R2, 3),
        ", p = ", ifelse(fire$`Pr(>F)` < 0.001, "<0.001", round(fire$`Pr(>F)`, 3)),
        "\nSite  F(", site$Df, ",", res_df, ") = ", round(site$F, 2),
        ", R\u00b2 = ", round(site$R2, 3),
        ", p = ", ifelse(site$`Pr(>F)` < 0.001, "<0.001", round(site$`Pr(>F)`, 3))
    )
}

perm_label <- make_perm_label(perm_base, perm_res_df)
# ── 3. Rebuild CAP scores with severity as a visual variable ─
scrs_sites_full <- cbind(wide_myco_soil, scrs_site)

# Severity as a continuous colour gradient on the ordination
p_site <- scrs_sites_full %>%
    ggplot(aes(x = CAP1, y = CAP2)) +
    geom_vline(xintercept = 0, color = "grey70", linetype = 2) +
    geom_hline(yintercept = 0, color = "grey70", linetype = 2) +
    geom_point(
        aes(
            colour = Fire_Treatment,
            shape = Fire_Treatment
        ),
        size = 6,
        stroke = 2
    ) +
    annotate(
        "text",
        x = -Inf, y = Inf,
        label = perm_label,
        hjust = -0.05, vjust = 1.3,
        size = 5
    ) +
    scale_colour_manual(
        values = c("B" = "darkred", "U" = "gold"),
        labels = c("B" = "Burnt", "U" = "Unburnt"),
        name = "Fire treatment"
    ) +
    scale_shape_manual(
        values = c("B" = 16, "U" = 1),
        labels = c("B" = "Burnt", "U" = "Unburnt"),
        name   = "Fire treatment"
    ) +
    labs(
        x = paste0("CAP1 (", proportions[1], "%)"),
        y = paste0("CAP2 (", proportions[2], "%)")
    ) +
    theme_classic() +
    theme(
        axis.text.x      = element_text(size = 14, colour = "black"),
        axis.text.y      = element_text(size = 14, colour = "black"),
        axis.title       = element_text(size = 16),
        legend.position  = "top",
        legend.text      = element_text(size = 12),
        legend.title     = element_text(size = 13)
    )

print(p_site)
ggsave("Plots/p_site.png", plot = p_site, width = 10, height = 7.5, bg = "white")
```

# Addresses reviewer comment: Line 143
# Tests whether AM and EcM fungi respond differently to fire
# RAREFACTION SECTION: Compare rarefied vs. non-rarefied data

```{r am_vs_ecm_alpha_rarefy_vs_unrarefy, fig.width=10, fig.height=7.5}
# import soil seq data
MER_Soil <- readRDS("Processed_data/Seq_dat/MER_Soil.rds") %>%
    mutate(
        Plot = str_replace(Plot, "(U|B)$", "") %>%
            str_replace("^NSSO", "NSS")
    ) %>%
    # remove empty cols
    select(-sample, qbit_DNA_conc_ng_uL)

# convert to wide format
temp <- MER_Soil %>%
    select(Plot, OTU, count) %>%
    pivot_wider(names_from = OTU, values_from = count, values_fill = 0)

mat <- temp %>%
    remove_rownames() %>%
    column_to_rownames("Plot")

# Step 1: Check read depth distribution across all samples
plot_read_totals_all <- rowSums(mat)
cat("Summary of total reads per plot (all OTUs):\n")
print(summary(plot_read_totals_all))
cat("\nPlots with fewer than 3,000 reads (will be dropped):\n")
print(plot_read_totals_all[plot_read_totals_all < 3000])

# Step 2: Rarefy to 3,000 reads
set.seed(42)
mat_rare <- rrarefy(
    mat[plot_read_totals_all >= 3000, ],
    sample = 3000
)

# Step 3: Convert rarefied matrix back to long format matching MER_Soil structure
MER_Soil_rare <- mat_rare %>%
    as.data.frame() %>%
    rownames_to_column("Plot") %>%
    pivot_longer(
        cols      = -Plot,
        names_to  = "OTU",
        values_to = "count"
    ) %>%
    filter(count > 0) %>% # drop true zeros
    left_join( # re-attach all OTU metadata
        MER_Soil %>%
            select(-count) %>%
            distinct(),
        by = c("Plot", "OTU")
    ) %>%
    mutate(Plot = as.factor(Plot))

# Step 4: Construct wide_myco_soil_rare by filtering for mycorrhizal taxa
Soil_Myco_rare <- MER_Soil_rare %>%
    filter(str_detect(guild, "mycorrhizal") | str_detect(guild, "Mycorrhizal"))

wide_myco_soil_rare <- Soil_Myco_rare %>%
    pivot_wider(
        names_from  = OTU,
        values_from = count,
        id_cols     = Plot,
        values_fill = 0
    ) %>%
    mutate(Plot = as.factor(Plot)) %>%
    left_join(
        All_Sites %>%
            select(
                Site, Plot, Fire_Treatment
            )
    )

# Step 5: Sanity checks
cat("\nPlots retained after rarefaction:", nrow(mat_rare), "\n")
cat("Plots in non-rarefied wide_myco:  ", nrow(wide_myco_soil), "\n")
cat("Plots in rarefied wide_myco_soil_rare:", nrow(wide_myco_soil_rare), "\n")
cat("OTUs in non-rarefied wide_myco:   ", ncol(wide_myco_soil) - 1, "\n")
cat("OTUs in rarefied wide_myco_soil_rare:", ncol(wide_myco_soil_rare) - 1, "\n")

# Step 6: Save
# write_rds(MER_Soil_rare,       "Processed_data/Seq_dat/Soil/MER_Soil_rare.rds")
# write_rds(wide_myco_soil_rare, "Processed_data/Seq_dat/Soil/wide_myco_soil_rare.rds")

# ============================================================
# Alpha Diversity and Venn Diagrams by Mycorrhizal Guild
# ============================================================

# ── 0. Additional Libraries ──────────────────────────────────
# (tidyverse, vegan, lme4, car, ggpubr, patchwork are loaded in setup)
library(fossil)
library(broom.mixed)
library(ggvenn)

# (wide_myco_soil, myco_tax_Soil, and All_Sites are already loaded and pre-joined in the previous chunk)

# ── 1. Identify OTU columns by guild ─────────────────────────
# Uses guild2 which contains 'EctoMycorrhizal' and 'Arbuscular Mycorrhizal'

ecm_otus <- myco_tax_Soil %>%
    filter(str_detect(guild2, regex("ectomycorrhizal", ignore_case = TRUE))) %>%
    pull(OTU)

am_otus <- myco_tax_Soil %>%
    filter(str_detect(guild2, regex("arbuscular", ignore_case = TRUE))) %>%
    pull(OTU)

# Subset to OTU columns present in the wide table
all_otu_cols <- wide_myco_soil %>%
    select(starts_with("ITSall")) %>%
    colnames()

ecm_cols <- intersect(paste0(ecm_otus), all_otu_cols)
am_cols <- intersect(paste0(am_otus), all_otu_cols)

cat("EcM OTUs retained:", length(ecm_cols), "\n")
cat("AM OTUs retained:", length(am_cols), "\n")

# ── 2. Create guild-specific community matrices ───────────────
meta_cols <- c("Site", "Plot", "Fire_Treatment")

wide_ecm <- wide_myco_soil %>%
    select(all_of(meta_cols), all_of(ecm_cols)) %>%
    filter(!if_all(all_of(ecm_cols), ~ . == 0))

wide_am <- wide_myco_soil %>%
    select(all_of(meta_cols), all_of(am_cols)) %>%
    filter(!if_all(all_of(am_cols), ~ . == 0))

# ── 2b. Guild-specific PERMANOVA + CAP ───────────────────────

# Helper: build a CAP plot with PERMANOVA annotation
make_cap_plot <- function(wide_df, otu_cols, perm_tbl, title_label, tag_label,
                          col_burnt, col_unburnt) {
    mat <- wide_df %>% select(all_of(otu_cols))
    res_df <- perm_tbl["Residual", "Df"]
    label <- make_perm_label(perm_tbl, res_df)

    cap <- capscale(mat ~ Fire_Treatment + Site,
        data     = wide_df,
        distance = "robust.aitchison",
        add      = TRUE
    )
    props <- round(cap$CCA$eig / cap$tot.chi * 100, 1)

    scrs_df <- cbind(wide_df, as.data.frame(scores(cap, display = "sites")))

    ggplot(scrs_df, aes(x = CAP1, y = CAP2)) +
        geom_vline(xintercept = 0, color = "grey70", linetype = 2) +
        geom_hline(yintercept = 0, color = "grey70", linetype = 2) +
        geom_point(aes(colour = Fire_Treatment, shape = Fire_Treatment),
            size = 6, stroke = 2
        ) +
        annotate("text",
            x = -Inf, y = Inf, label = label,
            hjust = -0.05, vjust = 1.3, size = 4
        ) +
        scale_colour_manual(
            values = c("B" = col_burnt, "U" = col_unburnt),
            labels = c("B" = "Burnt", "U" = "Unburnt"),
            name   = "Fire treatment"
        ) +
        scale_shape_manual(
            values = c("B" = 16, "U" = 1),
            labels = c("B" = "Burnt", "U" = "Unburnt"),
            name   = "Fire treatment"
        ) +
        labs(
            tag = tag_label,
            title = title_label,
            x = paste0("CAP1 (", props[1], "%)"),
            y = paste0("CAP2 (", props[2], "%)")
        ) +
        theme_classic() +
        theme(
            axis.text.x      = element_text(size = 13, colour = "black"),
            axis.text.y      = element_text(size = 13, colour = "black"),
            axis.title       = element_text(size = 14),
            legend.position  = "top",
            legend.text      = element_text(size = 11),
            legend.title     = element_text(size = 12),
            plot.title       = element_text(size = 14, face = "bold", hjust = 0.5),
            plot.tag         = element_text(size = 14, face = "bold")
        )
}

# EcM PERMANOVA
perm_ecm <- adonis2(
    wide_ecm %>% select(all_of(ecm_cols)) ~ Fire_Treatment + Site,
    data     = wide_ecm,
    distance = "robust.aitchison",
    by       = "margin"
)
print(perm_ecm)

# AM PERMANOVA
perm_am <- adonis2(
    wide_am %>% select(all_of(am_cols)) ~ Fire_Treatment + Site,
    data     = wide_am,
    distance = "robust.aitchison",
    by       = "margin"
)
print(perm_am)

# Build guild-specific CAP plots
p_cap_ecm <- make_cap_plot(
    wide_ecm, ecm_cols, perm_ecm,
    title_label = "EcM fungi",
    tag_label = "a)",
    col_burnt = "darkred",
    col_unburnt = "gold"
)

p_cap_am <- make_cap_plot(
    wide_am, am_cols, perm_am,
    title_label = "AM fungi",
    tag_label = "b)",
    col_burnt = "darkred",
    col_unburnt = "gold"
)

# Combined guild CAP figure
cap_combined <- p_cap_ecm + p_cap_am +
    plot_layout(guides = "collect") &
    theme(legend.position = "top")

print(cap_combined)
ggsave("Plots/cap_guild_combined.png",
    plot = cap_combined,
    width = 14, height = 7, bg = "white"
)

# ── 3. Alpha diversity function ───────────────────────────────
calc_alpha <- function(wide_df, otu_cols, guild_label) {
    wide_df %>%
        rowwise() %>%
        mutate(
            Shannon  = diversity(c_across(all_of(otu_cols)), index = "shannon"),
            Simpson  = diversity(c_across(all_of(otu_cols)), index = "simpson"),
            Chao1    = chao1(c_across(all_of(otu_cols))),
            Observed = sum(c_across(all_of(otu_cols))),
            Pielou   = ifelse(Observed > 1, Shannon / log(Observed), NA),
            Guild    = guild_label
        ) %>%
        ungroup()
}

alpha_ecm <- calc_alpha(wide_ecm, ecm_cols, "EcM")
alpha_am <- calc_alpha(wide_am, am_cols, "AM")

# ── 4. Fit LMMs by guild ──────────────────────────────────────
metrics <- c("Shannon", "Simpson", "Chao1", "Pielou")

fit_anova <- function(alpha_df, guild_label) {
    alpha_df <- alpha_df %>%
        mutate(Fire_Treatment = factor(Fire_Treatment, levels = c("U", "B")))

    map_dfr(metrics, function(metric) {
        formula <- as.formula(paste0(metric, " ~ Fire_Treatment + (1 | Site)"))
        model <- lmer(formula, data = alpha_df)
        anov <- Anova(model, test = "F")
        tidy_m <- tidy(model, effects = "fixed") %>%
            filter(term == "Fire_TreatmentB")

        tibble(
            Guild      = guild_label,
            Metric     = metric,
            Estimate   = tidy_m$estimate,
            Std_Error  = tidy_m$std.error,
            DF_num     = anov$Df,
            DF_denom   = anov$Df.res,
            F_value    = anov$F,
            P_value    = anov$`Pr(>F)`[1]
        )
    })
}

anova_ecm <- fit_anova(alpha_ecm, "EcM")
anova_am <- fit_anova(alpha_am, "AM")
anova_guild <- bind_rows(anova_ecm, anova_am)
print(anova_guild)
```

# ── 4b. Addresses reviewer comment: richness by Unburnt / Low / Medium / High ──
# Fire severity (continuous) is already in alpha_ecm / alpha_am via wide_myco_soil.
# For unburnt plots (Fire_Treatment == "U") set fire_severity to 0 so all plots
# sit on the same ordinal scale: 0 = Unburnt, 1 = Low, 2 = Medium, 3 = High.

```{r alpha_severity_lmm, fig.width = 12, fig.height = 6}
# ── Helper: join fire_severity from soil_guild_plot, recode unburnt to ──────
sev_lookup <- soil_guild_plot %>%
    select(Site, Plot, fire_severity) %>%
    distinct()

add_severity_cat <- function(alpha_df) {
    alpha_df %>%
        left_join(sev_lookup, by = c("Site", "Plot")) %>%
        mutate(
            # unburnt plots have no severity score; assign 0
            fire_severity_num = if_else(Fire_Treatment == "U", 0,
                as.numeric(fire_severity)
            )
        )
}

alpha_ecm_sev <- add_severity_cat(alpha_ecm)
alpha_am_sev <- add_severity_cat(alpha_am)

# ── LMM: Chao1 ~ fire_severity_num + (1 | Site) — continuous predictor ───────
fit_severity_lmm <- function(alpha_df, guild_label) {
    map_dfr(metrics, function(metric) {
        formula <- as.formula(paste0(metric, " ~ fire_severity_num + (1 | Site)"))
        model <- lmer(formula, data = alpha_df)
        anov <- Anova(model, test = "F")

        tibble(
            Guild    = guild_label,
            Metric   = metric,
            DF_num   = anov$Df[1],
            DF_denom = anov$Df.res[1],
            F_value  = anov$F[1],
            P_value  = anov$`Pr(>F)`[1]
        )
    })
}

sev_lmm_ecm <- fit_severity_lmm(alpha_ecm_sev, "EcM")
sev_lmm_am <- fit_severity_lmm(alpha_am_sev, "AM")
sev_lmm_all <- bind_rows(sev_lmm_ecm, sev_lmm_am)
print(sev_lmm_all)

# Continuous gradient colours: Unburnt (grey) → Low → Moderate → High → Very High → Extreme (dark red)
severity_gradient_cols <- c("#cccccc", "#fee08b", "#f4a261", "#e76f51", "#c0392b", "#7b0000")

plot_severity <- function(alpha_df, lmm_df, guild_label, index, tag_label) {
    p_val <- lmm_df %>%
        filter(Guild == guild_label, Metric == index) %>%
        pull(P_value)

    p_label <- case_when(
        p_val <= 0.001 ~ "p <0.001",
        p_val <= 0.01 ~ "p <0.01",
        p_val <= 0.05 ~ paste0("p = ", round(p_val, 3)),
        p_val <= 0.1 ~ paste0("p = ", round(p_val, 2)),
        TRUE ~ "ns"
    )

    # Fit continuous LMM
    model <- lmer(as.formula(paste0(index, " ~ fire_severity_num + (1 | Site)")),
        data = alpha_df
    )

    # Build prediction grid for regression ribbon
    sev_range <- seq(
        min(alpha_df$fire_severity_num, na.rm = TRUE),
        max(alpha_df$fire_severity_num, na.rm = TRUE),
        length.out = 100
    )
    pred_df <- data.frame(fire_severity_num = sev_range)
    mm <- model.matrix(~fire_severity_num, data = pred_df)
    pred_df$fit <- as.vector(mm %*% fixef(model))
    pred_df$se <- sqrt(diag(mm %*% as.matrix(vcov(model)) %*% t(mm)))
    pred_df <- pred_df %>%
        mutate(lwr = fit - 1.96 * se, upr = fit + 1.96 * se)

    y_pos <- max(alpha_df[[index]], na.rm = TRUE) * 1.08

    ggplot() +
        geom_ribbon(
            data = pred_df,
            aes(x = fire_severity_num, ymin = lwr, ymax = upr),
            fill = "grey70", alpha = 0.4
        ) +
        geom_line(
            data = pred_df,
            aes(x = fire_severity_num, y = fit),
            linewidth = 1.2, color = "black"
        ) +
        geom_jitter(
            data = alpha_df,
            aes(x = fire_severity_num, y = .data[[index]], colour = fire_severity_num),
            width = 0.08, alpha = 0.6, size = 2.5
        ) +
        scale_colour_gradientn(
            colours = severity_gradient_cols,
            limits  = c(0, 5),
            breaks  = c(0, 1, 2, 3, 4, 5),
            labels  = c("Unburnt", "Low", "Moderate", "High", "Very High", "Extreme"),
            name    = "Severity"
        ) +
        scale_x_continuous(
            breaks = c(0, 1, 2, 3, 4, 5),
            labels = c("Unburnt", "Low", "Moderate", "High", "Very High", "Extreme")
        ) +
        annotate("text",
            x = 0, y = y_pos,
            label = p_label, hjust = 0, size = 6
        ) +
        labs(
            tag = tag_label,
            x   = "Fire Severity",
            y   = paste0(index, " (", guild_label, ")")
        ) +
        theme_classic() +
        theme(
            legend.position  = "none",
            axis.text.x      = element_text(hjust = 0.5, size = 16, face = "bold"),
            axis.text.y      = element_text(size = 16, face = "bold"),
            axis.title.x     = element_text(size = 18, face = "bold"),
            axis.title.y     = element_text(size = 18, face = "bold"),
            axis.line        = element_line(linewidth = 2, colour = "black"),
            plot.tag         = element_text(size = 16, face = "bold")
        )
}

p_sev_ecm_chao1 <- plot_severity(alpha_ecm_sev, sev_lmm_all, "EcM", "Chao1", "a)")
p_sev_am_chao1 <- plot_severity(alpha_am_sev, sev_lmm_all, "AM", "Chao1", "b)")

combined_sev_chao1 <- p_sev_ecm_chao1 + p_sev_am_chao1
print(combined_sev_chao1)
ggsave("Plots/chao1_by_severity_guild.png",
    plot = combined_sev_chao1, width = 14, height = 7, bg = "white"
)

p_sev_ecm_shannon <- plot_severity(alpha_ecm_sev, sev_lmm_all, "EcM", "Shannon", "a)")
p_sev_am_shannon <- plot_severity(alpha_am_sev, sev_lmm_all, "AM", "Shannon", "b)")

combined_sev_shannon <- p_sev_ecm_shannon + p_sev_am_shannon
print(combined_sev_shannon)
ggsave("Plots/shannon_by_severity_guild.png",
    plot = combined_sev_shannon, width = 14, height = 7, bg = "white"
)


# ── 5. Plot Chao1 by guild ────────────────────────────────────
plot_chao1 <- function(alpha_df, anova_df, guild_label,
                       fill_burnt, fill_unburnt, tag_label) {
    p_val <- anova_df %>%
        filter(Guild == guild_label, Metric == "Chao1") %>%
        pull(P_value)

    p_label <- case_when(
        p_val <= 0.001 ~ "p <0.001",
        p_val <= 0.01 ~ "p <0.01",
        p_val <= 0.05 ~ paste0("p = ", round(p_val, 3)),
        p_val <= 0.1 ~ paste0("p = ", round(p_val, 2)),
        TRUE ~ "ns"
    )

    # Fit LMM to extract Estimated Marginal Means
    m_df <- alpha_df %>%
        mutate(Fire_Treatment = factor(Fire_Treatment, levels = c("U", "B")))

    model <- lmer(Chao1 ~ Fire_Treatment + (1 | Site), data = m_df)

    emms <- as.data.frame(emmeans::emmeans(model, ~Fire_Treatment)) %>%
        mutate(
            Fire_Treatment = case_when(
                Fire_Treatment == "U" ~ "Unburnt",
                Fire_Treatment == "B" ~ "Burnt"
            ),
            Fire_Treatment = factor(Fire_Treatment, levels = c("Unburnt", "Burnt"))
        )

    raw_df <- alpha_df %>%
        mutate(
            Fire_Treatment = case_when(
                Fire_Treatment == "U" ~ "Unburnt",
                Fire_Treatment == "B" ~ "Burnt"
            ),
            Fire_Treatment = factor(Fire_Treatment,
                levels = c("Unburnt", "Burnt")
            )
        )

    y_pos <- max(raw_df$Chao1, na.rm = TRUE) * 1.05

    ggplot() +
        geom_jitter(
            data = raw_df, aes(x = Fire_Treatment, y = Chao1),
            width = 0.2, alpha = 0.3, size = 2, color = "grey40"
        ) +
        geom_pointrange(
            data = emms, aes(
                x = Fire_Treatment, y = emmean,
                ymin = lower.CL, ymax = upper.CL,
                fill = Fire_Treatment
            ),
            shape = 21, size = 1.2, stroke = 1.5, color = "black"
        ) +
        scale_fill_manual(values = c(
            Unburnt = fill_unburnt,
            Burnt = fill_burnt
        )) +
        annotate("text",
            x = 0.7, y = y_pos,
            label = p_label, hjust = 0.5, size = 6
        ) +
        labs(
            tag   = tag_label,
            x     = "Fire Treatment",
            y     = paste0("Chao1 Estimated Richness (", guild_label, ")")
        ) +
        theme_classic() +
        theme(
            legend.position  = "none",
            axis.text.x      = element_text(hjust = 0.5, size = 16, face = "bold"),
            axis.text.y      = element_text(size = 16, face = "bold"),
            axis.title.x     = element_text(size = 18, face = "bold"),
            axis.title.y     = element_text(size = 18, face = "bold"),
            axis.line        = element_line(linewidth = 2, colour = "black"),
            plot.tag         = element_text(size = 16, face = "bold")
        )
}

p_chao1_ecm <- plot_chao1(alpha_ecm, anova_guild, "EcM",
    fill_burnt   = "#2F4F4F",
    fill_unburnt = "#cccccc",
    tag_label    = "a)"
)

p_chao1_am <- plot_chao1(alpha_am, anova_guild, "AM",
    fill_burnt   = "#7B3F00",
    fill_unburnt = "#D2B48C",
    tag_label    = "b)"
)

plotly::ggplotly(p_chao1_ecm + p_chao1_am)
# ── 6. Venn diagrams by guild ─────────────────────────────────
get_otu_sets_guild <- function(df, otu_cols) {
    list(
        Unburnt = df %>%
            filter(Fire_Treatment == "U") %>%
            select(all_of(otu_cols)) %>%
            summarise(across(everything(), ~ any(. > 0))) %>%
            pivot_longer(everything(),
                names_to = "OTU", values_to = "present"
            ) %>%
            filter(present) %>% pull(OTU),
        Burnt = df %>%
            filter(Fire_Treatment == "B") %>%
            select(all_of(otu_cols)) %>%
            summarise(across(everything(), ~ any(. > 0))) %>%
            pivot_longer(everything(),
                names_to = "OTU", values_to = "present"
            ) %>%
            filter(present) %>% pull(OTU)
    )
}

otu_sets_ecm <- get_otu_sets_guild(wide_ecm, ecm_cols)
otu_sets_am <- get_otu_sets_guild(wide_am, am_cols)

venn_ecm <- ggvenn(
    otu_sets_ecm,
    fill_color      = c("#cccccc", "#2F4F4F"),
    show_percentage = TRUE,
    digits          = 0,
    set_name_size   = 6,
    stroke_size     = 2,
    text_size       = 7
) +
    labs(tag = "c)", title = "EcM fungi") +
    theme(
        plot.tag   = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )

venn_am <- ggvenn(
    otu_sets_am,
    fill_color      = c("#D2B48C", "#7B3F00"),
    show_percentage = TRUE,
    digits          = 0,
    set_name_size   = 6,
    stroke_size     = 2,
    text_size       = 7
) +
    labs(tag = "d)", title = "AM fungi") +
    theme(
        plot.tag   = element_text(size = 16, face = "bold"),
        plot.title = element_text(size = 16, face = "bold", hjust = 0.5)
    )

combined_guild <- (p_chao1_ecm + p_chao1_am) / (venn_ecm + venn_am)

# Display the final plot block
print(combined_guild)

# Save the plot
ggsave("Plots/combined_guild.png", plot = combined_guild, width = 12, height = 10, bg = "white")
```



# Myc (Hyphal) Samples {.tabset}

## Box Plots by Severity

```{r hyph-box-severity, fig.width=12, fig.height=8}
hyph_guild_plot %>%
    filter(!is.na(severity)) %>%
    ggplot(aes(x = severity, y = RA_plot, fill = as.factor(severity))) +
    geom_boxplot(alpha = 0.7, outlier.shape = 21) +
    geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
    facet_wrap(~guild2, scales = "free_y") +
    # scale_fill_manual(values = severity_colours) +
    labs(
        title = "Myc Samples: Guild Relative Abundance by Fire Severity",
        x = "Fire Severity",
        y = "Relative Abundance (%)",
        fill = "Severity"
    )
```

## LMM Statistics

```{r hyph-lmm-stats}
hyph_stats <- hyph_guild_plot %>%
    filter(!is.na(severity)) %>%
    group_by(guild2) %>%
    nest() %>%
    mutate(
        model = map(data, lmer(RA_plot ~ severity + (1 | Site), data = .x))
    ) %>%
    filter(!map_lgl(anova, is.null)) %>%
    mutate(
        F_value = map_dbl(anova, ~ .x$`F value`[1]),
        p_value = map_dbl(anova, ~ .x$`Pr(>F)`[1])
    ) %>%
    select(guild2, F_value, p_value) %>%
    arrange(p_value) %>%
    mutate(
        significance = case_when(
            p_value < 0.001 ~ "***",
            p_value < 0.01 ~ "**",
            p_value < 0.05 ~ "*",
            p_value < 0.1 ~ ".",
            TRUE ~ "ns"
        )
    )

knitr::kable(hyph_stats,
    digits = 4,
    caption = "Myc Samples: LMM ANOVA results for guild ~ severity + (1|Site)"
)
```

## Stacked Bar by Site

```{r hyph-stacked-site, fig.width=14, fig.height=8}
hyph_guild_plot %>%
    filter(!is.na(severity)) %>%
    mutate(Site = fct_reorder(Site, as.numeric(factor(severity)))) %>%
    ggplot(aes(x = Plot, y = RA_plot, fill = guild2)) +
    geom_bar(stat = "identity") +
    facet_wrap(~severity, scales = "free_x") +
    labs(
        title = "Myc Samples: Guild Composition by Fire Severity",
        x = "Plot",
        y = "Relative Abundance (%)",
        fill = "Guild"
    ) +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```


# Soil ITS Samples {.tabset}

## Box Plots by Severity

```{r soil-box-severity, fig.width=12, fig.height=8}
soil_guild_plot %>%
    filter(!is.na(severity)) %>%
    ggplot(aes(x = severity, y = RA_plot, fill = severity)) +
    geom_boxplot(alpha = 0.7, outlier.shape = 21) +
    geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
    facet_wrap(~guild2, scales = "free_y") +
    scale_fill_manual(values = severity_colours) +
    labs(
        title = "Soil Samples: Guild Relative Abundance by Fire Severity",
        x = "Fire Severity",
        y = "Relative Abundance (%)",
        fill = "Severity"
    ) +
    theme_pub
```

## LMM Statistics

```{r soil-lmm-stats}
soil_stats <- soil_guild_plot %>%
    filter(!is.na(severity)) %>%
    group_by(guild2) %>%
    nest() %>%
    mutate(
        model = map(data, possibly(~ {
            lmer(RA_plot ~ severity + (1 | Site), data = .x)
        }, otherwise = NULL)),
        anova = map(model, possibly(~ {
            as.data.frame(anova(.x))
        }, otherwise = NULL))
    ) %>%
    filter(!map_lgl(anova, is.null)) %>%
    mutate(
        F_value = map_dbl(anova, ~ .x$`F value`[1]),
        p_value = map_dbl(anova, ~ .x$`Pr(>F)`[1])
    ) %>%
    select(guild2, F_value, p_value) %>%
    arrange(p_value) %>%
    mutate(
        significance = case_when(
            p_value < 0.001 ~ "***",
            p_value < 0.01 ~ "**",
            p_value < 0.05 ~ "*",
            p_value < 0.1 ~ ".",
            TRUE ~ "ns"
        )
    )

knitr::kable(soil_stats,
    digits = 4,
    caption = "Soil Samples: LMM ANOVA results for guild ~ severity + (1|Site)"
)
```

## Stacked Bar by Site

```{r soil-stacked-site, fig.width=14, fig.height=8}
soil_guild_plot %>%
    filter(!is.na(severity)) %>%
    mutate(Site = fct_reorder(Site, as.numeric(factor(severity)))) %>%
    ggplot(aes(x = Plot, y = RA_plot, fill = guild2)) +
    geom_bar(stat = "identity") +
    facet_wrap(~severity, scales = "free_x") +
    labs(
        title = "Soil Samples: Guild Composition by Fire Severity",
        x = "Plot",
        y = "Relative Abundance (%)",
        fill = "Guild"
    ) +
    theme_pub +
    theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 7))
```


# Comparison: Myc vs Soil {.tabset}

## Side-by-Side Box Plots

```{r combined-boxplots, fig.width=14, fig.height=10}
# Combine the two datasets
combined <- bind_rows(
    hyph_guild_plot %>%
        filter(!is.na(severity)) %>%
        mutate(Sample_type = "Myc (Hyphal)"),
    soil_guild_plot %>%
        filter(!is.na(severity)) %>%
        mutate(Sample_type = "Soil ITS")
)

combined %>%
    ggplot(aes(x = severity, y = RA_plot, fill = severity)) +
    geom_boxplot(alpha = 0.7, outlier.shape = 21) +
    geom_jitter(width = 0.15, alpha = 0.3, size = 1) +
    facet_grid(Sample_type ~ guild2, scales = "free_y") +
    scale_fill_manual(values = severity_colours) +
    labs(
        title = "Comparison of Guild Relative Abundance Across Fire Severities",
        subtitle = "Myc (Hyphal) vs Soil ITS Samples",
        x = "Fire Severity",
        y = "Relative Abundance (%)",
        fill = "Severity"
    ) +
    theme(strip.text.y = element_text(angle = 0))
```

## Summary Statistics Table

```{r combined-summary}
combined_summary <- combined %>%
    group_by(Sample_type, guild2, severity) %>%
    summarise(
        n = n_distinct(Plot),
        mean_RA = mean(RA_plot, na.rm = TRUE),
        sd_RA = sd(RA_plot, na.rm = TRUE),
        median_RA = median(RA_plot, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    arrange(Sample_type, guild2, severity)

knitr::kable(combined_summary,
    digits = 2,
    caption = "Summary statistics of guild relative abundance by sample type and severity"
)
```

## Combined LMM Results

```{r combined-lmm}
combined_stats <- bind_rows(
    hyph_stats %>% mutate(Sample_type = "Myc (Hyphal)"),
    soil_stats %>% mutate(Sample_type = "Soil ITS")
) %>%
    select(Sample_type, guild2, F_value, p_value, significance) %>%
    arrange(Sample_type, p_value)

knitr::kable(combined_stats,
    digits = 4,
    caption = "Combined LMM ANOVA results: guild ~ severity + (1|Site)"
)
```

# Session Info

```{r session-info}
sessionInfo()
```
